Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-06-03 docs/supplementary_material.tex
Incorporated all draft3 comments and added new figures.

% Document
\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{lipsum}

@@

\section*{Appendix 1: Implementation of the Crofton Method}

The algorithm for fitting the Crofton Method \citep{Crofton1971a} proceeds as follows. First, obtain a dataset
with $n$ hosts where each host has some parasite intensity 0 to $p_{max}$. Starting
with the full dataset, guess a vector of pre-mortality parameters $(N_p, \mu_p, k_p)$ where $N_p$ is the total number of hosts before mortality, $\mu_p$ is the parasite intensity before mortality, and $k_p$ is the parasite aggregation before mortality. Given
these parameters, use a negative binomial distribution to calculate the predicted number of hosts with $0, 1, 2, \dots
p_{max}$ parasites.  Compare the expected number of hosts with $0, 1, 2, \dots  p_{max}$ parasites
to the observed number hosts with $0, 1, 2, \dots p_{max}$ parasites and calculate
the $\chi^2$-squared statistic associated with your observed and predicted vectors. In reality, one often has to bin the parasite intensity data because all
parasite intensities are not represented in the dataset. Continue to guess $(N_p,
mu_p, k_p)$ vectors until a set of parameters is found that minimizes the $\chi^2$-
squared statistic.

Second, choose a truncation value ($t_1$) such that $t_1 <
p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t_1$ and repeat the above
iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
that minimizes the $\chi^2$-squared statistic on the truncated data. Choose a new
truncated value $t_2 < t_1$ and repeat the first two steps. Continue to truncate
the dataset until it only contains hosts with 0, 1, and 2 parasites (or 3 bins).
Because the method attempts to estimate three parameters, at least 3 classes are
needed for all 3 parameters to be identifiable \citep{Royce1990}.  Parasite-induced host mortality is traditionally identified by then plotting your truncation values $t_i$ against your different values of $N_{po}$ and looking for a distinct kink in the resulting plot.

@@

\section*{Appendix 2: Implementation of the Adjei Method}

The Adjei Method for estimating PIHM has two steps \citep{Adjei1986}.  The first
step is to estimate the parameters of the pre-mortality host-parasite
distribution using the Crofton Method.  The three parameters estimated are the
total number of hosts before mortality $N_p$,  the mean number of parasites per
host before mortality $\mu_p$, and the aggregation of parasites before
mortality given by the parameter $k_p$ from a negative binomial distribution.
When $k_p$ is small, parasites are highly aggregated among hosts and when
$k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.  The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and above]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 4.

The second step of the Adjei Method is to estimate the shape of the host survival function. \cite{Adjei1986} assume that the host survival function follows the logistic form

\begin{equation}
    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
parameters of the logistic function. Generally, a larger $a$ allows for hosts to
tolerate larger parasite intensities before experiencing parasite-induced mortality
and a larger $b$ leads to a more rapid decline in the probability of host
survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
experience mortality.

To estimate this function, the Adjei Method first
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \mu_p, k_p)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

% \section*{Appendix 3: Implementation of the Likelihood Method}

% As described in the main text, the probability mass function for the likelihood method is given by

% \begin{equation}
%     P(x | \text{survival}) = \dfrac{P(\text{survival} | x) * P(x)}{P(\text{survival})}
%     \label{eq:concept}
% \end{equation}

% One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$. Therefore equation \ref{eq:concept} can be written as

% \begin{equation}
%     P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
%     \label{eq:dist}
% \end{equation}

% Using this equation, we seek to find estimates for $\mu_p$, $k_p$, $a$, and $b$
% that maximize the likelihood of equation \ref{eq:dist} given a host-parasite
% dataset. The first approach is use a standard maximization algorithm to try to
% find all four parameters.  However, for small number of hosts and/or a gradual
% survival function, the gradient of the likelihood will not contain much
% information (check this) and the maximization will likely fail or give
% unrealistic results. An alternative approach would be to parameterize $\mu_p$
% and $k_p$ using the Crofton Method and then use equation \ref{eq:dist} to find the values for $a$ and $b$ that maximize the likelihood of a host-parasite dataset.

\section*{Appendix 3: Additional Figures}

@@

\section*{Appendix 4: Code and unit tests for estimating parasite-induced host mortality}

Python code, unit tests, and a help file for the Crofton Method, the Adjei Method and the Likelihood Method can be found in \verb*#pihm_methods.py#, \verb*#test_pihm_methods.py#, and \verb*#help_file.txt#, respectively.

>>>>

% Document
\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{lipsum}
\usepackage{hyperref}

@@

\section*{SI 1: Implementation of the Crofton Method}

The algorithm for fitting the Crofton Method \citep{Crofton1971a} proceeds as follows. First, obtain a dataset
with $n$ hosts where each host has some parasite intensity 0 to $p_{max}$. Starting
with the full dataset, guess a vector of pre-mortality parameters $(N_{p1}, \mu_{p1}, k_{p1})$ where $N_{p1}$ is the total number of hosts before mortality, $\mu_{p1}$ is the parasite intensity before mortality, and $k_{p1}$ is the parasite aggregation before mortality. Given
these parameters, use a negative binomial distribution to calculate the predicted number of hosts with $0, 1, 2, \dots
p_{max}$ parasites.  Compare the expected number of hosts with $0, 1, 2, \dots  p_{max}$ parasites
to the observed number hosts with $0, 1, 2, \dots p_{max}$ parasites and calculate
the $\chi^2$-squared statistic associated with the observed and predicted vectors. In reality, one often has to bin the parasite intensity data because all
parasite intensities are not represented in the dataset. Continue to guess $(N_{p1},
\mu_{p1}, k_{p1})$ vectors until a set of parameters is found that minimizes the $\chi^2$-
squared statistic.

Second, choose a truncation value ($t_2$) such that $t_2 <
p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t_2$ and repeat the above
iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
that minimizes the $\chi^2$-squared statistic on the truncated data. Choose a new
truncated value $t_3 < t_2$ and repeat the first two steps. Continue to truncate
the dataset until it only contains hosts with 0, 1, and 2 parasites (or 3 bins).
Because the method attempts to estimate three parameters, at least 3 classes are
needed for all 3 parameters to be identifiable \citep{Royce1990}.

Once the iterative procedure has been completed, parasite-induced host mortality is traditionally identified by plotting the different truncation values $t_{i}$ against the different values of $N_{pi}$ and looking for a distinct ``kink'' in the resulting plot.   Once the ``kink'' as occurred, the values of $N_{pi}$ will typically remain close to constant as $t_i$ is decreased further.  The ``true'' pre-mortality parameters $N_{pt}$, $\mu_{pt}$, and $k_{pt}$ are taken to be at the point where the ``kink'' occurs.

@@

\section*{SI 2: Implementation of the Adjei Method}

The Adjei Method for estimating PIHM has two steps \citep{Adjei1986}.  The first
step is to estimate the parameters of the pre-mortality host-parasite
distribution using the Crofton Method (see \emph{SI} 1).  The three parameters estimated are the
total number of hosts before mortality $N_p$,  the mean number of parasites per
host before mortality $\mu_p$, and the aggregation of parasites before
mortality given by the parameter $k_p$ from a negative binomial distribution.
When $k_p$ is small, parasites are highly aggregated among hosts and when
$k_p$ is large parasites are more evenly distributed across hosts \citep{Wilson2002}.  The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in \emph{SI} 4]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 4.

The second step of the Adjei Method is to make the assumption that infection, host mortality, and sampling occur in that order and are temporally \cite{Adjei1986}. Next, the \citeauthor{Adjei1986} assume that the host survival function follows the logistic form

\begin{equation}
    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
parameters of the logistic function. Generally, a larger $a$ allows for hosts to
tolerate larger parasite intensities before experiencing parasite-induced mortality
and a larger $b$ leads to a more rapid decline in the probability of host
survival as parasite intensity increases. The value $\exp(a / b)$ is referred
to as the $LD_{50}$. Individuals with loads higher than this will have a greater than 50\% chance of death.

By taking the first and second derivatives of equation \ref{eq:logistic}, one can easily find that the maximum rate of decline in host survival probability with increasing parasite intensity occurs at the $LD_{50}$ and has a value of $b / 4$.  This is in many ways analogous to the parasite pathogenicity parameter $\alpha$ given in classic macroparasite models, which specifies the slope of the linear relationship between between host death rate and parasite intensity \citep{AndersonandMay1978, Isham1995}.  The parameter $a$ is easily interpreted by holding $b$ constant and looking at how a one unit change in $a$ affects the log parasite intensity at which some percentage $p$ of hosts experience mortality.  Letting $a_1$ and $a_2$ be two different values of $a$ and $x_1$ and $x_2$ be two different parasite intensities, a bit of rearranging of equation \ref{eq:logistic} gives.

\begin{align*}
    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
    - \log{\frac{p}{1 - p}} &= a_1 - b \log{x_2}\\
    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
\end{align*}
If $a_2 - a_1 = 1$, than the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.

To estimate the parameters in equation \ref{eq:logistic}, the Adjei Method first
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \mu_p, k_p)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

\section*{SI 3: Additional Figures}

@@

\section*{SI 4: Code and unit tests for estimating parasite-induced host mortality}

Python code, unit tests, and a help file for the Crofton Method, the Adjei Method and the Likelihood Method can be found at \url{https://github.com/mqwilber/parasite_mortality}
